Contrasting snow algae communities in the mountains of British Columbia, Canada

### 2019-12-21 manuscript/manuscript.Rmd
minor changes title and abstract

---
title: "Contrasting alpine and subalpine snow algae communities in British Columbia, Canada"
# who grows in the snow? multi-hued snow algae communities
# different coloured alpine and subalpine snows harbour constrasting snow algae communities
# contrasting snow algae communities in di
# not just dirty snow; alpine snow algae communities have different assemblages
# green, orange, or red snow caused by distinct
# snow algae communities causing colored snow in alpine and subalpine summer snow 
# distinct snow algae communities across an elevational gradient
author:
  - name: Casey B. Engstrom
    institute: SFU
  - name: Kurt M. Yakimovich
    institute: SFU
  - name: Lynne M. Quarmby
    institute: SFU
    email: quarmby@sfu.ca
institute:
  - SFU: Dept. Molecular Biology and Biochemistry, Simon Fraser University, 8888 University Drive, Burnaby, BC, V5A 1S6, Canada
output: 
  bookdown::word_document2:
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=templates/scholarly-metadata.lua'
      - '--lua-filter=templates/author-info-blocks.lua'
    reference_docx: templates/word-template.docx
  pdf_document: 
    fig_caption: yes
    number_sections: yes
csl: templates/nature-communications.csl
bibliography: casey1.bib
abstract: |
  Snow algae grow in melting summer snowfields throughout polar and alpine regions, forming extensive blooms that reduce albedo and increase snow melt. Green, orange, and red patches of snow algae can harbour morphologically diverse communities, but despite their global ubiquity the community composition of snow algae blooms are poorly understood. We Illumina sequenced 18S and *rbcL* amplicons from 33 snow algae samples collected from a range of alpine and subalpine habitats in the Coast Range of British Columbia, Canada. Communities were dominated by algae closely related to *Sanguina* and *Chloromonas*, the latter including genetically distinct sub-groups including clusters matching *Chlainomonas* and several sub-groups that could represent potentially novel clades. Above treeline snow algae communities were dominated by *Sanguina*, while *Chloromonas krienitzii* was found in snow runnels near treeline, which transitioned from green to orange throughout the season. *Chlainomonas* was abundant in samples across all elevations. Our results suggest that while snow algae diversity at the genus level is relatively low much remains to be explored at the species level, and that communities of snow algae can differ greatly from patch to patch of snow. 
---

@@

Snow algae are cold-adapted microalgae that bloom in melting summer snow throughout alpine and polar regions [@Leya2013]. Red snow algae blooms can cover vast areas of snowfield, which decreases snow surface albedo---by one estimate, enough to substantially increase the rate of snow melt [@Ganey2017]. Earlier snow melt exposes darker surfaces below, contributing to snow-albedo feedback and global warming [@Dery2007], which can threaten water supplies [@Staudinger2014] and worsen drought [@Barnhart2016]. Thus, snow algae could play an important role in snow melt regulation, but little is known about how these blooms form. We don't know how cells colonize the white snow to initiate the bloom, what triggers snow algae to divide and produce pigment, or why one patch of snow turns red while the neighboring snow remains white. Critically, we lack basic understanding of the species composition and diversity present in snow algae blooms, which is fundamental to resolving long-standing mysteries about snow algae bloom formation.

>>>>

---
title: "Contrasting snow algae communities in the mountains of British Columbia, Canada"
# who grows in the snow? multi-hued snow algae communities
# different coloured alpine and subalpine snows harbour constrasting snow algae communities
# contrasting snow algae communities in di
# not just dirty snow; alpine snow algae communities have different assemblages
# green, orange, or red snow caused by distinct
# snow algae communities causing colored snow in alpine and subalpine summer snow 
# distinct snow algae communities across an elevational gradient
author:
  - name: Casey B. Engstrom
    institute: SFU
  - name: Kurt M. Yakimovich
    institute: SFU
  - name: Lynne M. Quarmby
    institute: SFU
    email: quarmby@sfu.ca
institute:
  - SFU: Dept. Molecular Biology and Biochemistry, Simon Fraser University, 8888 University Drive, Burnaby, BC, V5A 1S6, Canada
output: 
  bookdown::word_document2:
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=templates/scholarly-metadata.lua'
      - '--lua-filter=templates/author-info-blocks.lua'
    reference_docx: templates/word-template.docx
  pdf_document: 
    fig_caption: yes
    number_sections: yes
csl: templates/nature-communications.csl
bibliography: casey1.bib
abstract: |
  Snow algae grow in melting summer snowfields throughout polar and alpine regions, forming extensive blooms that reduce albedo and increase snow melt. Green, orange, and red patches of snow algae can harbour morphologically diverse communities, but despite their global ubiquity the community composition of snow algae blooms are poorly understood. We Illumina sequenced 18S and *rbcL* amplicons from 33 snow algae samples collected from a range of alpine and subalpine habitats in the Coast Range of British Columbia, Canada. Communities were dominated by algae closely related to *Sanguina* and *Chloromonas*, the latter including genetically distinct sub-groups of *Chlainomonas* and several potentially novel clades. Above tree line communities were dominated by *Sanguina*, and below tree line by *Chloromonas*. *Chlainomonas* was abundant in samples across all elevations. We also observed blooms of *Chloromonas krienitzii* transition from green to orange throughout the season. We found snow algae community composition was highly variable from patch to patch, even on the same mountain. Although relatively few genera dominated snow algae communities, we found high diversity at the species level. 
---

@@

Snow algae are cold-adapted microalgae that bloom on melting summer snow throughout alpine and polar regions. Snow algae darken the snow surface and decrease albedo[@lutz_biogeography_2016;onuma_temporal_2019]---by one estimate, enough to substantially increase snow melt[@ganey_role_2017]. Estimates of global coverage are lacking, but snow algae blooms can cover many square kilometers of glacier summer snow[@ganey_role_2017], are present on snow overlying sea ice[@gradinger_snow_1996], and are reported from every continent[@segawa_bipolar_2018;@duval_snow_1999;@marchant_snow_1982;@yoshimura_community_1997;@vimercati_nieves_2019].

Sanger seq isolates

Sanger seq field samples

18S HTS

<!-- re work this to follow the flow of my presentation -->
Microscopy reveals a diversity of algae in patches of green, orange, or pink snow (Fig. 1). Amplicon high-throughput sequencing (HTS) shows that most snow algae are typically Chlamydomonadales, although other orders are occasionally reported[@remias_two_2019;@lutz_integrated_2015;@soto_microbial_2020]. Three genera are frequently found in colored snow. *Sanguina* are red or orange cells found across the Arctic, which was recently re-assigned to its own genus based on Sanger sequencing of field samples[@prochazkova_sanguina_2019].  *Chloromonas* currently has 35 species listed on GenBank, 13 of which are snow algae[@matsuzaki_taxonomic_2019]. Microscopy suggests *Chloromonas* is prevalent in forested sites[@hoham_distribution_1979,@nedbalova_ecology_2008], but recent 18S metabarcoding found *Chloromonas* was also widespread in Arctic red snow in low relative abundance[@lutz_biogeography_2016]. The third genus *Chlainomonas* looks completely different from *Chloromonas*, but *rbcL* phylogeny suggests the two are closely related[@novis_two_2008]. Both *Chlainomonas* and *Sanguina* are colored red by astaxanthin, but *Chlainomonas* cells are typically twice the diameter, with four flagella and a distinct cell wall[@remias_ecophysiology_2016] (Fig 2x). *Chlainomonas* and *Sanguina* remain uncultured[@prochazkova_ecophysiological_2018;@novis_two_2008].

Microscopic identification is complicated by the multiple life stages typical of microalgae: the same species can look completely different between life stages; conversely, the same species can look nearly identical. The freshwater algae *Haematococcus pluvialis* undergoes mitosis as green cells, but stressors such as high light and nutrient starvation triggers formation of a dormant cyst with thick cell walls and astaxanthin pigment[@ota_carotenoid_2018;@kobayashi_light-independent_1997]. Attempts to induce cyst development in snow algae remain unsuccessful; the bright pigments and ornamented cell walls of *Chloromonas* observed in the field bear little resemblance to the green cells in culture[@leya_response_2009]. Amplicon studies have found that green snow is caused by different species than red snow[@lutz_integrated_2015;@terashima_microbial_2017;@soto_microbial_2020], so it remains unknown whether red snow species undergo a green life phase at all.

Snow algae are the primary producer in a little-studied microbiome that includes protists, fungi, bacteria, and viruses[@naff_phylogeny_2013;@lutz_biogeography_2016;@hamilton_primary_2017;@terashima_microbial_2017;@hisakawa_metagenomic_2015]. Despite nearly a century of microscopic study of red snow[@leya_snow_2013], how snow algae blooms develop remains a mystery. How do cells propagate in this harsh environment? What limits the extent of snow algae blooms? Critically, we lack basic understanding of snow algae diversity and life history.

<!-- this from draft 1 of casey1, above is from MBB806 -->
 Earlier snow melt exposes darker surfaces below, contributing to snow-albedo feedback and global warming [@Dery2007], which can threaten water supplies [@Staudinger2014] and worsen drought [@Barnhart2016]. Thus, snow algae could play an important role in snow melt regulation, but little is known about how these blooms form. We don't know how cells colonize the white snow to initiate the bloom, what triggers snow algae to divide and produce pigment, or why one patch of snow turns red while the neighboring snow remains white. Critically, we lack basic understanding of the species composition and diversity present in snow algae blooms, which is fundamental to resolving long-standing mysteries about snow algae bloom formation.
